### 2021-06-10 papers/thesis/full/ch3-disk-slices/disk-slices.tex
Added index of notation. Uses makeindex.

We will show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations on the interval \cite{olver2013fast}, and its analogues on the disk \cite{vasil2016tensor} and triangle \cite{olver2018recurrence,olver2019triangle}. On the disk-slice the family of weights we consider are of the form
\bseqn{
	\Wabc(x,y) = (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c
}
for $\alpha \leq x \leq \beta,$ $-\rho(x) \leq y \leq \rho(x)$. The corresponding OPs denoted $\hdopnkabc(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
\bseqn{
	\hdopnkabc(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
}
where $C_{n,k} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{n-j} y^j$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs we will see in \bsrefdef{def:ds:OPconstruction}.

@@

We can construct 2D orthogonal polynomials on $\Omega$ from 1D orthogonal polynomials on the intervals \([\alpha,\beta]\) and \([\gamma,\delta]\).

@@

For disk slices and trapeziums, we specialise \bsrefprop{prop:ds:construction} in the following definitions. First we introduce notation for two families of univariate OPs.
\begin{definition}\label{def:ds:OPconstruction}
	Let $\genjacw^{(a,b,c)}(x)$ and $\jacw^{(a,b)}(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\bseqn{
	\begin{cases}
		\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
		\jacw^{(a,b)}(x) &:= (\delta-x)^a \: (x - \gamma)^b
	\end{cases}
}
and define the associated inner products by:
\bseqnnumber{
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ds:ipgenjac} \\
	\ip<p, q>_{\jacw^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: \jacw^{(a,b)}(y)\: \D y \label{eqn:ds:ipjac}
}
where
\bseqnnumber{
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: \jacw^{(a,b)}(y) \: \D y. \label{eqn:ds:ipnormalisations}
}
Denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in \bsrefeqn{eqn:ds:ipgenjac}, and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{\jac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in \bsrefeqn{eqn:ds:ipjac}.
\end{definition}

We can now write down our 2D OP family.
\begin{definition}\label{def:ds:constuction}
	For parameters $a,b,c,d \in \R$, define the four-parameter 2D orthogonal polynomials as the set $\{\hdopnk^{(a,b,c,d)}\}$ where:
\bseqn{
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(d,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
}
\end{definition}

@@

$\{\hdopnk^{(a,b,c,d)}\}$ are orthogonal with respect to the weight
\bseqn{
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: \jacw^{(d,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
}
assuming that either \bsrefcondition{cond:trap} holds, or \bsrefcondition{cond:disk} holds with $\jacw^{(a,b)}$ being an even function (i.e. $a = b$, and we can hence denote the weight as $\jacw^{(a)}(x) = w^{(a,a)}_P(x) = (\delta-x^2)^a$). That is,
\bseqn{
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{W^{(a,b,c,d)}} &= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j}
}
where for $f, g : \Omega \to \R$ the inner product is defined as 
\bseqn{
	\ip< f, \: g >_{W^{(a,b,c,d)}} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x.
}

We can see that they are indeed orthogonal using the change of variable $t = \frac{y}{\rho(x)}$, for the following normalisation:
\bseqnnumber{
	&\ip< \hdopnkabcd, \: \hdopmjabcd >_\Wabcd \\
	&=  \iint_\Omega \: \Big[ \genjacnmk^{(a,b,c+d+2k+1)}(x) \: \genjacmmj^{(a,b,c+d+2j+1)}(x) \: \rho(x)^{k+j}  \nonumber \\
		&\quadfour \cdot \: \jac_k^{(d,c)}\fpr(\frac{y}{\rho(x)}) \: \jac_j^{(d,c)}\fpr(\frac{y}{\rho(x)})  \: \Wabcd(x,y) \Big] \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b,c+d+2k+1)}}(x) \: {\genjacmmj^{(a,b,c+d+2j+1)}}(x) \: \genjacw^{(a,b,c+d+k+j+1)}(x) \: \D x \Big) \nonumber \\
		&\quadfour \cdot \: \Big( \int_\gamma^\delta \: {\jac_k^{(d,c)}}(t) \:  {\jac_j^{(d,c)}}(t) \: \jacw^{(d,)}(t) \: \D t \Big) \nonumber \\
	&= \normjac^{(d,c)} \: \delta_{k,j} \: \int_\alpha^\beta \: {\genjacnmk^{(a,b,c+d+2k+1)}}(x) \: {\genjac_{m-k}^{(a,b,c+d+2k+1)}}(x) \: \genjacw^{(a,b,c+d+2k+1)}(x) \: \D x \nonumber \\ 
	&= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j}. \label{eqn:ds:normhdop}
}

@@

\begin{proof}
The 3-term recurrence for multiplication by $x$ follows from \bsrefeqn{eqn:ds:Hrecurrence}. For the recurrence for multiplication by $y$, since $\{\hdopmjabcd\}$ for $m = 0,\dots,n+1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n+1$ polynomial, we can expand $y \: \hdopnkabcd(x,y) = \sum_{m=0}^{n+1} \sum_{j=0}^m c_{m,j} \: \hdopmjabcd(x,y)$. These coefficients are given by
\bseqn{
	c_{m,j} = {\ip< y \: \hdopnkabcd, \hdopmjabcd >_{\Wabcd}}{\norm{\hdopmjabcd}^{-2}_{\Wabcd}}.
}
Recall from \bsrefeqn{eqn:ds:normhdop} that $\norm{\hdopmjabcd}_{\Wabcd}^2 = \normgenjac^{(a,b,c+d+2j+1)} \: \normjac^{(d,c)}$. Then for $m = 0,\dots,n+1$, $j = 0,\dots,m$, using the change of variable $t = \frac{y}{\rho(x)}$:
\bseqn{
	&\ip<y \hdopnkabcd, \hdopmjabcd>_\Wabcd \\
	&=  \iint_\Omega \hdopnkabcd(x,y) \: \hdopmjabcd(x,y) \: y \: \Wabcd(x,y) \: dy \: dx \\
	&= \Big( \int^\beta_\alpha \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \genjacmmj^{(a, b, c+d+2j+1)}(x) \: \rho(x)^{k+j+2} \: \genjacw^{(a, b,c+d)}(x) \: \D x \Big) \\
		&\quadfour \cdot \: \Big( \int^\delta_{\gamma} \jac_k^{(d,c)}(t) \: \jac_j^{(d,c)}(t) \: t \: \jacw^{(d,c)}(t) \: \D t \Big) \\
	&= \Big( \int^\beta_\alpha \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \genjacmmj^{(a, b, c+d+2j+1)}(x) \: \genjacw^{(a, b,c+d+k+j+2)}(x) \: \D x \Big) \\
		&\quadfour \cdot \: \Big( \int^\delta_{\gamma} \jac_k^{(d,c)}(t) \: \jac_j^{(d,c)}(t) \: t \: \jacw^{(d,c)}(t) \: \D t \Big) \\
	&=
	\begin{cases}
    		\delta_k^{(d,c)} \: \normjac^{(d,c)} \: \normgenjac^{(a, b, c+d+2k+3)} \: \ip<\genjacnmk^{(a, b, c+d+2k+1)}, \genjac_{m-k-1}^{(a, b, c+d+2k+3)}>_{\genjacw^{(a, b, c+d+2k+3)}} \\
			\quadeight \quadeight \quadeight \: \: \text{if } j = k+1 \\
		\gamma_k^{(d,c)} \: \normjac^{(d,c)} \: \normgenjac^{(a, b, c+d+2k+1)} \: \ip<\genjacnmk^{(a, b, c+d+2k+1)}, \: \rho(x) \genjac_{m-k}^{(a, b, c+d+2k+1)}>_{\genjacw^{(a, b, c+d+2k+1)}} \\
			\quadeight \quadeight \quadeight \: \: \text{if } j = k \\
		\delta_{k-1}^{(d,c)} \: \normjac^{(d,c)} \: \normgenjac^{(a, b, c+d+2k-1)} \: \ip<\genjacnmk^{(a, b, c+d+2k-1)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b, c+d+2k-1)}>_{\genjacw^{(a, b, c+d+2k-1)}} \\
			\quadeight \quadeight \quadeight \: \: \text{if } j = k-1 \\
		\quadeight \quadfour 0 \quadfour \quadeight \text{otherwise}
      	\end{cases}
}
where, by orthogonality,
\bseqn{
	\ip<\genjacnmk^{(a, b, c+d+2k+1)}, \genjac_{m-k-1}^{(a, b, c+d+2k+3)}>_{\genjacw^{(a, b, c+d+2k+3)}} &= 0 \qfor m < n-1, \\
	\ip<\genjacnmk^{(a, b, c+d+2k+1)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b, c+d+2k-1)}>_{\genjacw^{(a, b, c+d+2k-1)}} &= 0 \qfor m < n-1.
}

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: 
\bseqn{
\bighdopabcd_n := \begin{pmatrix}
		\hdopabcd_{n,0}(x,y) \\
		\vdots \\
		\hdopabcd_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
\quad \quad 
\bighdopabcd := \begin{pmatrix}
		\bighdopabcd_0 \\
		\bighdopabcd_1 \\
		\bighdopabcd_2 \\
		\vdots \\
	\end{pmatrix}
}
and set $J_x^{(a,b,c,d)}, J_y^{(a,b,c,d)}$ as the Jacobi matrices corresponding to
\bseqnnumber{
	J_x^{(a,b,c,d)} \: \bighdopabcd(x,y) &= x \: \bighdopabcd(x,y), \label{eqn:ds:jacobimatricesdefinitionx} \\
	J_y^{(a,b,c,d)} \: \bighdopabcd(x,y) &= y \: \bighdopabcd(x,y). \label{eqn:ds:jacobimatricesdefinitiony}
}
The matrices $J_x^{(a,b,c,d)}, J_y^{(a,b,c,d)}$ act on the coefficients vector of a function's expansion in the $\{\hdopnkabcd\}$ basis. For example, let $a, b$ be general parameters and a function $f(x,y)$ defined on $\Omega$ be approximated by its expansion $f(x,y) = \bighdopabcd(x,y)^\top \vec{f}$. Then $x \: f(x,y)$ is approximated by $\bighdopabcd(x,y)^\top {J_x^{(a,b,c,d)\top}} \vec{f}$. In other words, ${J_x^{(a,b,c,d)\top}} \vec{f}$ is the coefficients vector for the expansion of the function $(x,y) \mapsto x \: f(x,y)$ in the $\{\hdopnkabcd\}$ basis.

@@

In this section, we concentrate on the disk-slice case, and simply note that similar arguments apply for the trapezium case. Recall that, for the disk-slice,
\bseqn{
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
}
where
\bseqn{
	\begin{cases}
		(\alpha,\beta) &\subset (0,1) \\
		(\gamma,\delta) &:= (-1,1) \\
		\rho(x) &:= (1-x^2)^{\half}
	\end{cases}.
}
The 2D OPs on the disk-slice $\Omega$, orthogonal with respect to the weight
\bseqn{
	\Wabc(x,y) &:= \genjacw^{(a,b,2c)}(x) \: \jacw^{(c)}\fpr(\frac{y}{\rho(x)}) \\
	&= (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c, \quad (x,y) \in \Omega,
}
are then given by:
\bseqn{
	\hdopnkabc(x,y) := \genjacnmk^{(a, b, 2c+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(c,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega
}
where the 1D OPs $\{\genjac^{(a, b, c)}_n\}$ are orthonormal on the interval $(\alpha, \beta)$ with respect to the weight
\bseqn{
	\genjacw^{(a,b,c)}(x) :=  (\beta - x)^a \: (x - \alpha)^b \: \rho(x)^{c}
}
and the 1D OPs $\{\jac^{(c, c)}_n\}$ are orthonormal on the interval $(\gamma, \delta) = (-1, 1)$ with respect to the weight
\bseqn{
	\jacw^{(c)}(x) := (1-x)^c \: (1+x)^c = (1-x^2)^c.
}
Denote the weighted OPs by
\bseqn{
	\bigWabc(x,y) := \Wabc(x,y) \: \bighdopabc(x,y),
}
and recall that a function $f(x,y)$ defined on $\Omega$ is approximated by its expansion $f(x,y) = \bighdopabc(x,y)^\top \vec{f}$.

@@

Finally,
\bseqn{
	&\int_\alpha^\beta \: \genjacnmk^{(a, b, 2c+2k+1)} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} \: \rho^{k+j} \: \rho' \: \genjacw^{(a+1, b+1, 2c+2)} \: \D x \\
	&= \normgenjac^{(a, b, 2c+2k+1)} \: \ip< \genjacnmk^{(a, b, 2c+2k+1)}, \: \rho \: \rho' \: \genjacw^{(1, 1, j-k)} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} >_{\genjacw^{(a, b, 2c+2k+1)}}
}
which is also zero for $m < n - 3$. Thus
\bseqn{
	&\ip<\ppx{x} \hdopnkabc, \hdop^{(a+1, b+1, c+1)}_{m,j} >_{W^{(a+1, b+1, c+1)}} = 0
}
for $m < n - 3$, $j \notin \{k-2, k\}$, showing that the only possible non-zero coefficients $c_{m,j}^x$ are when $m = n-3,n-2,n-1$ and $j = k-2,k$ $(j \le m)$.

>>>>

We will show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations on the interval \cite{olver2013fast}, and its analogues on the disk \cite{vasil2016tensor} and triangle \cite{olver2018recurrence,olver2019triangle}. On the disk-slice the family of weights we consider are of the form \nomenclature[a]{$a,b,c,d$}{Lower case letters $a,b,c,d$ refer to OP family parameters (unless otherwise stated)}
\bseqn{
	\Wabc(x,y) = (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c
}
for $\alpha \leq x \leq \beta,$ $-\rho(x) \leq y \leq \rho(x)$. The corresponding OPs denoted $\hdopnkabc(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
\bseqn{
	\hdopnkabc(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
}
where $C_{n,k} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{n-j} y^j$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs we will see in \bsrefdef{def:ds:OPconstruction}.

@@

We can construct 2D orthogonal polynomials on $\Omega$ from 1D orthogonal polynomials on the intervals $[\alpha,\beta]$ and $[\gamma,\delta]$.

@@

For disk slices and trapeziums, we specialise \bsrefprop{prop:ds:construction} in the following definitions. First we introduce notation for two families of univariate OPs. \nomenclature[weight]{$\genjacw^{(a,b,c)}$, $\jacw^{(a,b)}$}{Weight functions for the one-dimensional OPs $\{\genjac_n^{(a,b,c)}\}$ and $\{\jac_n^{(a,b)}\}$ respectively} 
\begin{definition}\label{def:ds:OPconstruction}
	Let $\genjacw^{(a,b,c)}(x)$ and $\jacw^{(a,b)}(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\bseqn{
	\begin{cases}
		\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
		\jacw^{(a,b)}(x) &:= (\delta-x)^a \: (x - \gamma)^b
	\end{cases}
}
and define the associated inner products by: \nomenclature[omega]{$\normgenjac$, $\normjac$}{Normalising constants for inner products} \nomenclature[innerw]{$\ip< p, \: q >_w$}{A one-dimensional inner product with weight $w$}
\bseqnnumber{
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ds:ipgenjac} \\
	\ip<p, q>_{\jacw^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: \jacw^{(a,b)}(y)\: \D y \label{eqn:ds:ipjac}
}
where
\bseqnnumber{
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: \jacw^{(a,b)}(y) \: \D y. \label{eqn:ds:ipnormalisations}
}
Denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by \nomenclature[R]{$\genjac_n^{(a,b,c)}$}{Non-classical three-parameter univariate OP of degree $n$} $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in \bsrefeqn{eqn:ds:ipgenjac}, and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{\jac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in \bsrefeqn{eqn:ds:ipjac}.
\end{definition}

We can now write down our 2D OP family. \nomenclature[Hnk]{$\hdopnk^{(a,b,c,d)}$}{Four-parameter bivariate OP of degree $n$ and order $k$ defined on the given domain and by the given inner product}
\begin{definition}\label{def:ds:constuction}
	For parameters $a,b,c,d \in \R$, define the four-parameter 2D orthogonal polynomials as the set $\{\hdopnk^{(a,b,c,d)}\}$ where:
\bseqn{
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(d,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
}
\end{definition}

@@

$\{\hdopnk^{(a,b,c,d)}\}$ are orthogonal with respect to the weight \nomenclature[Wa]{$W^{(a)}, W^{(a,b,c)}, W^{(a,b,c,d)}$}{Weight function for relevant multidimensional OPs, with the given parameters}
\bseqn{
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: \jacw^{(d,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
}
assuming that either \bsrefcondition{cond:trap} holds, or \bsrefcondition{cond:disk} holds with $\jacw^{(a,b)}$ being an even function (i.e. $a = b$, and we can hence denote the weight as $\jacw^{(a)}(x) = w^{(a,a)}_P(x) = (\delta-x^2)^a$). That is,
\bseqn{
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{\hdop^{(a,b,c,d)}} &= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j}
}
where for $f, g : \Omega \to \R$ the inner product is defined as \nomenclature[innerH]{$\ip< p, \: q >_{\hdopabcd}$}{A multidimensional inner product over $\Omega$ with the weight $\Wabcd$ associated with the OPs $\{\hdopnkabcd\}$}
\bseqn{
	\ip< f, \: g >_{\hdop^{(a,b,c,d)}} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x.
}

We can see that they are indeed orthogonal using the change of variable $t = \frac{y}{\rho(x)}$, for the following normalisation:
\bseqnnumber{
	&\ip< \hdopnkabcd, \: \hdopmjabcd >_{\hdopabcd} \\
	&=  \iint_\Omega \: \Big[ \genjacnmk^{(a,b,c+d+2k+1)}(x) \: \genjacmmj^{(a,b,c+d+2j+1)}(x) \: \rho(x)^{k+j}  \nonumber \\
		&\quadfour \cdot \: \jac_k^{(d,c)}\fpr(\frac{y}{\rho(x)}) \: \jac_j^{(d,c)}\fpr(\frac{y}{\rho(x)})  \: \Wabcd(x,y) \Big] \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b,c+d+2k+1)}}(x) \: {\genjacmmj^{(a,b,c+d+2j+1)}}(x) \: \genjacw^{(a,b,c+d+k+j+1)}(x) \: \D x \Big) \nonumber \\
		&\quadfour \cdot \: \Big( \int_\gamma^\delta \: {\jac_k^{(d,c)}}(t) \:  {\jac_j^{(d,c)}}(t) \: \jacw^{(d,)}(t) \: \D t \Big) \nonumber \\
	&= \normjac^{(d,c)} \: \delta_{k,j} \: \int_\alpha^\beta \: {\genjacnmk^{(a,b,c+d+2k+1)}}(x) \: {\genjac_{m-k}^{(a,b,c+d+2k+1)}}(x) \: \genjacw^{(a,b,c+d+2k+1)}(x) \: \D x \nonumber \\ 
	&= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j}. \label{eqn:ds:normhdop}
}

@@

\begin{proof}
The 3-term recurrence for multiplication by $x$ follows from \bsrefeqn{eqn:ds:Hrecurrence}. For the recurrence for multiplication by $y$, since $\{\hdopmjabcd\}$ for $m = 0,\dots,n+1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n+1$ polynomial, we can expand $y \: \hdopnkabcd(x,y) = \sum_{m=0}^{n+1} \sum_{j=0}^m c_{m,j} \: \hdopmjabcd(x,y)$. These coefficients are given by \nomenclature[abs]{$\abs{m}$}{The absolute value of a scalar $m$} \nomenclature[norm]{$\norm{f}$}{The norm of a function f, defined by the square root of the relevant inner product} \nomenclature[normH]{$\norm{f}_{\hdopabcd}$}{The norm of a function f, defined by the square root of the inner product over $\Omega$ given by the weight $\Wabcd$ associated with the OPs $\{\hdopnkabcd\}$}
\bseqn{
	c_{m,j} = {\ip< y \: \hdopnkabcd, \hdopmjabcd >_{\hdopabcd}}{\norm{\hdopmjabcd}^{-2}_{\hdopabcd}}.
}
Recall from \bsrefeqn{eqn:ds:normhdop} that $\norm{\hdopmjabcd}_{\hdopabcd}^2 = \normgenjac^{(a,b,c+d+2j+1)} \: \normjac^{(d,c)}$. Then for $m = 0,\dots,n+1$, $j = 0,\dots,m$, using the change of variable $t = \frac{y}{\rho(x)}$:
\bseqn{
	&\ip<y \hdopnkabcd, \hdopmjabcd>_{\hdopabcd} \\
	&=  \iint_\Omega \hdopnkabcd(x,y) \: \hdopmjabcd(x,y) \: y \: \Wabcd(x,y) \: dy \: dx \\
	&= \Big( \int^\beta_\alpha \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \genjacmmj^{(a, b, c+d+2j+1)}(x) \: \rho(x)^{k+j+2} \: \genjacw^{(a, b,c+d)}(x) \: \D x \Big) \\
		&\quadfour \cdot \: \Big( \int^\delta_{\gamma} \jac_k^{(d,c)}(t) \: \jac_j^{(d,c)}(t) \: t \: \jacw^{(d,c)}(t) \: \D t \Big) \\
	&= \Big( \int^\beta_\alpha \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \genjacmmj^{(a, b, c+d+2j+1)}(x) \: \genjacw^{(a, b,c+d+k+j+2)}(x) \: \D x \Big) \\
		&\quadfour \cdot \: \Big( \int^\delta_{\gamma} \jac_k^{(d,c)}(t) \: \jac_j^{(d,c)}(t) \: t \: \jacw^{(d,c)}(t) \: \D t \Big) \\
	&=
	\begin{cases}
    		\delta_k^{(d,c)} \: \normjac^{(d,c)} \: \normgenjac^{(a, b, c+d+2k+3)} \: \ip<\genjacnmk^{(a, b, c+d+2k+1)}, \genjac_{m-k-1}^{(a, b, c+d+2k+3)}>_{\genjacw^{(a, b, c+d+2k+3)}} \\
			\quadeight \quadeight \quadeight \: \: \text{if } j = k+1 \\
		\gamma_k^{(d,c)} \: \normjac^{(d,c)} \: \normgenjac^{(a, b, c+d+2k+1)} \: \ip<\genjacnmk^{(a, b, c+d+2k+1)}, \: \rho(x) \genjac_{m-k}^{(a, b, c+d+2k+1)}>_{\genjacw^{(a, b, c+d+2k+1)}} \\
			\quadeight \quadeight \quadeight \: \: \text{if } j = k \\
		\delta_{k-1}^{(d,c)} \: \normjac^{(d,c)} \: \normgenjac^{(a, b, c+d+2k-1)} \: \ip<\genjacnmk^{(a, b, c+d+2k-1)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b, c+d+2k-1)}>_{\genjacw^{(a, b, c+d+2k-1)}} \\
			\quadeight \quadeight \quadeight \: \: \text{if } j = k-1 \\
		\quadeight \quadfour 0 \quadfour \quadeight \text{otherwise}
      	\end{cases}
}
where, by orthogonality,
\bseqn{
	\ip<\genjacnmk^{(a, b, c+d+2k+1)}, \genjac_{m-k-1}^{(a, b, c+d+2k+3)}>_{\genjacw^{(a, b, c+d+2k+3)}} &= 0 \qfor m < n-1, \\
	\ip<\genjacnmk^{(a, b, c+d+2k+1)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b, c+d+2k-1)}>_{\genjacw^{(a, b, c+d+2k-1)}} &= 0 \qfor m < n-1.
}

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: \nomenclature[Hvectorabcd]{$\bighdopabcd$}{Vector of the disk-slice/trapezium OPs ordered by degree, with parameters $a,b,c,d$} \nomenclature[Hvectornabcd]{$\bighdopabcd_n$}{(Sub-)Vector of the disk-slice/trapezium OPs of degree $n$, with parameters $a,b,c,d$}
\bseqn{
	\bighdopabcd_n := 
		\begin{pmatrix}
			\hdopabcd_{n,0}(x,y) \\
			\vdots \\
			\hdopabcd_{n,n}(x,y)
		\end{pmatrix} \in \R^{n+1}, 
	\quad \quad 
	\bighdopabcd := 
		\begin{pmatrix}
			\bighdopabcd_0 \\
			\bighdopabcd_1 \\
			\bighdopabcd_2 \\
			\vdots \\
		\end{pmatrix}
}
and set $J_x^{(a,b,c,d)}, J_y^{(a,b,c,d)}$ as the Jacobi matrices corresponding to
\bseqnnumber{
	J_x^{(a,b,c,d)} \: \bighdopabcd(x,y) &= x \: \bighdopabcd(x,y), \label{eqn:ds:jacobimatricesdefinitionx} \\
	J_y^{(a,b,c,d)} \: \bighdopabcd(x,y) &= y \: \bighdopabcd(x,y). \label{eqn:ds:jacobimatricesdefinitiony}
}
The matrices $J_x^{(a,b,c,d)}, J_y^{(a,b,c,d)}$ act on the coefficients vector of a function's expansion in the $\{\hdopnkabcd\}$ basis. For example, let $a, b$ be general parameters and a function $f(x,y)$ defined on $\Omega$ be approximated by its expansion $f(x,y) = \bighdopabcd(x,y)^\top \vec{f}$. Then $x \: f(x,y)$ is approximated by $\bighdopabcd(x,y)^\top {J_x^{(a,b,c,d)\top}} \vec{f}$. In other words, ${J_x^{(a,b,c,d)\top}} \vec{f}$ is the coefficients vector for the expansion of the function $(x,y) \mapsto x \: f(x,y)$ in the $\{\hdopnkabcd\}$ basis.

@@

In this section, we concentrate on the disk-slice case, and simply note that similar arguments apply for the trapezium case. Recall that, for the disk-slice,
\bseqn{
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
}
where
\bseqn{
	\begin{cases}
		(\alpha,\beta) &\subset (0,1) \\
		(\gamma,\delta) &:= (-1,1) \\
		\rho(x) &:= (1-x^2)^{\half}
	\end{cases}.
}
The 2D OPs on the disk-slice $\Omega$, orthogonal with respect to the weight
\bseqn{
	\Wabc(x,y) &:= \genjacw^{(a,b,2c)}(x) \: \jacw^{(c)}\fpr(\frac{y}{\rho(x)}) \\
	&= (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c, \quad (x,y) \in \Omega,
}
are then given by:
\bseqn{
	\hdopnkabc(x,y) := \genjacnmk^{(a, b, 2c+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(c,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega
}
where the 1D OPs $\{\genjac^{(a, b, c)}_n\}$ are orthonormal on the interval $(\alpha, \beta)$ with respect to the weight
\bseqn{
	\genjacw^{(a,b,c)}(x) :=  (\beta - x)^a \: (x - \alpha)^b \: \rho(x)^{c}
}
and the 1D OPs $\{\jac^{(c, c)}_n\}$ are orthonormal on the interval $(\gamma, \delta) = (-1, 1)$ with respect to the weight
\bseqn{
	\jacw^{(c)}(x) := (1-x)^c \: (1+x)^c = (1-x^2)^c.
}
Denote the weighted OPs by \nomenclature[Wvectorabcd]{$\bigWabcd$}{Vector of the weighted disk-slice/trapezium OPs ordered by degree, with parameters $a,b,c,d$}
\bseqn{
	\bigWabc(x,y) := \Wabc(x,y) \: \bighdopabc(x,y),
}
and recall that a function $f(x,y)$ defined on $\Omega$ is approximated by its expansion $f(x,y) = \bighdopabc(x,y)^\top \vec{f}$.

@@

Finally,
\bseqn{
	&\int_\alpha^\beta \: \genjacnmk^{(a, b, 2c+2k+1)} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} \: \rho^{k+j} \: \rho' \: \genjacw^{(a+1, b+1, 2c+2)} \: \D x \\
	&= \normgenjac^{(a, b, 2c+2k+1)} \: \ip< \genjacnmk^{(a, b, 2c+2k+1)}, \: \rho \: \rho' \: \genjacw^{(1, 1, j-k)} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} >_{\genjacw^{(a, b, 2c+2k+1)}}
}
which is also zero for $m < n - 3$. Thus
\bseqn{
	&\ip<\ppx{x} \hdopnkabc, \hdop^{(a+1, b+1, c+1)}_{m,j} >_{\hdop^{(a+1, b+1, c+1)}} = 0
}
for $m < n - 3$, $j \notin \{k-2, k\}$, showing that the only possible non-zero coefficients $c_{m,j}^x$ are when $m = n-3,n-2,n-1$ and $j = k-2,k$ $(j \le m)$.
